Supermassive black hole and host bulge affairs \\ II. The red and blue sequence in the $M_{\rm BH} - M_{\rm *,sph}$ diagram.

### 2015-05-04 papers/MbhMsph/mm.tex
mon

\begin{abstract}
We present correlations between black hole mass, $M_{\rm BH}$, 
and both the host galaxy's and spheroid's infrared luminosity, $L_{\rm gal}$ and $L_{\rm sph}$, 
and the spheroid's stellar mass, $M_{\rm *,sph}$,
built upon a high-quality data set.
Our sample, the largest ever used, counts 66 local galaxies with a dynamical measurement of $M_{\rm BH}$.
%Luminosities come from high signal-to-noise $3.6\rm~\mu m$ \emph{Spitzer} satellite imagery, 
%which is minimally affected by dust and star formation. 
%These luminosities were derived from our state-of-the-art multicomponent galaxy decompositions, 
%that take into account bulges, disks, bars, spiral arms, rings, haloes, extended or unresolved nuclear sources and partially depleted cores, 
%and that -- for the first time -- were checked to be consistent with the galaxy kinematics. 
Spheroid luminosities come from high signal-to-noise $3.6\rm~\mu m$ \emph{Spitzer} satellite imagery, 
and were derived from our state-of-the-art {\bf many-component} galaxy decompositions 
that, for the first time, were checked to be consistent with the galaxy kinematics. 
{\bf limit of 250 words, no room for mentioning the gaalxy components, more appropriate for data paper}
Previous studies have used galaxy samples that were overwhelmingly dominated by high-mass, early-type objects.
Instead, our sample includes 17 spiral galaxies, half of which have $M_{\rm BH} < 10^7\rm~M_\odot$, 
and allows us to better investigate the poorly studied low-mass end of the $M_{\rm BH} - M_{\rm *,sph}$ correlation.
%We provide updated linear regressions 
%(symmetrical, to be compared with theoretical expectations, and non-symmetrical, to be used to predict black hole masses) 
%for S\'ersic/core-S\'ersic galaxies and for different morphological types (E/S0, Sp). 
The bulges of early-type (E + S0) galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$, 
consistent with a dry-merging formation scenario,
and define a tight \emph{red sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$.
On the other hand, the bulges of late-type (Sp) galaxies define a much steeper \emph{blue sequence}, 
with $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently than the host bulge. 
We additionally report that: i) S\'ersic galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.48 \pm 0.20}$, a less steep sequence than previously reported; 
ii) bulges with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
are not offset to lower $M_{\rm BH}$ from the correlation defined by bulges with $n_{\rm sph}>2$; 
iii) $L_{\rm sph}$ and $L_{\rm gal}$ correlate equally well with $M_{\rm BH}$, in terms of intrinsic scatter, only for early-type galaxies; 
once reasonable numbers of spiral galaxies are included, the correlation with $L_{\rm sph}$ is better than that with $L_{\rm gal}$.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\columnwidth]{images/mbh_vs_mag_tot.pdf}
\caption{Black hole mass plotted against $3.6\rm~\mu m$ galaxy absolute magnitude. 
Symbols are coded according to the galaxy morphological type: red circle = E, red star = E/S0, 
red upward triangle = S0, blue downward triangle = S0/Sp, blue square = Sp, black diamond = merger. 
Empty symbols represent core-S\'ersic spheroids, whereas filled symbols are used for S\'ersic spheroids. 
Four spiral galaxies had their magnitudes overestimated and are shown as upper limits. 
The red dashed line indicates the BCES bisector linear regression for 45 early-type galaxies (E+S0), 
with the red shaded area denoting its $1\sigma$ uncertainty. 
We were not able to obtain any meaningful linear regression for the late-type (Sp) galaxies. }
\label{fig:mbhmaggal}
\end{center}
\end{figure}

@@

\subsection{Pseudo- versus classical bulges}
Current views distinguish between classical bulges, which are considered to be spheroidal, pressure-supported systems, 
formed through violent processes, such as hierarchical clustering via minor mergers, 
and pseudo-bulges, thought to be disk-like, rotation-supported systems, 
built from secular evolution processes, such as instabilities of their surrounding disk or bar. 
Pseudo-bulges are notoriously hard to identify \citep{graham2013review,graham2014review,graham2015pseudo,graham2015review}.
For example, mergers can create bulges that rotate (e.g.~\citealt{bekki2010,keselmannusser2012}), 
and bars can spin-up classical bulges (e.g.~\citealt{saha2012}), 
thus rotation is not a definitive signature of a pseudobulge. 
Furthermore, many galaxies host both a pseudo- and a classical bulge (e.g.~\citealt{erwin2003,erwin2015,athanassoula2005,Gadotti2009,
macarthur2009,dosanjosdasilva2013,seidel2015}). 
In the recent literature, pseudo- and classical bulges have been frequently told apart by means of a divide at a 
S\'ersic index of $n_{\rm sph}=2$ (e.g.~\citealt{sani2011,beifiori2012}), 
although, from a selection of hundreds of disc galaxies imaged in the $K$-band, 
\cite{grahamworley2008} observed no such bimodality in the bulge S\'ersic indices. 
While pseudo-bulges are expected to have exponential-like surface brightness profiles ($n_{\rm sph} \simeq 1$), 
being disky components that formed from their surrounding exponential disks 
(e.g.~\citealt{bardeen1975,hohl1975,combessanders1981,combes1990,pfennigerfriedli1991}), 
it has been shown that mergers can create bulges with $n_{\rm sph}<2$
(e.g.~\citealt{elichemoral2011,scannapieco2011,querejeta2015}), 
just as low-luminosity elliptical galaxies (not built from the secular evolution of a disk)
are well known to have $n_{\rm sph}<2$ and even $n_{\rm sph}<1$ (e.g.~\citealt{davies1988,youngcurrie1994}). \\
\cite{sani2011} reported that pseudo-bulges -- which they labelled as such according to the $n_{\rm sph}<2$ criterion -- 
with low black hole masses ($M_{\rm BH} < 10^7\rm~M_\odot$) are significantly displaced from the correlation 
traced by classical bulges. 
In Figure \ref{fig:pseudob}, we show the distribution of spheroid S\'ersic indices\footnote{The spheroid S\'ersic index 
is taken from our one-dimensional fit of the galaxy's major-axis surface brightness profile (\emph{Paper I}).} 
in the $M_{\rm BH} - L_{\rm sph}$ diagram.
Our aim is to check whether $n_{\rm sph}<2$ bulges 
are offset to lower black hole masses from the correlation defined by $n_{\rm sph}>2$ bulges. 
To do this, we fit a symmetrical linear regression to the bulges that have $n_{\rm sph}>2$ 
and we compute the vertical offset of all bulges from the regression. 
In the inset of Figure \ref{fig:pseudob}, we plot the vertical offset against $n_{\rm sph}$. 
Among the 23 bulges with $n_{\rm sph}<2$, 12 have positive vertical offset and 11 have negative vertical offset. 
\cite{kormendy2015review} provides a list of many pseudo-bulges classification criteria, including the divide at $n_{\rm sph}=2$, 
and cautions that each individual criterion has a failure rate of 0-25\% \citep{fisherdrory2015}. 
If this is true, we should have that no less than 75\% of $n_{\rm sph}<2$ bulges display negative vertical offset. 
What we observe, instead, is that there are the same number of $n_{\rm sph}<2$ bulges lying above and below 
the correlation defined by $n_{\rm sph}>2$ bulges\footnote{One 
reaches the same conclusion when using the vertical offset from the correlation defined by bulges with $n_{\rm sph}>3$. 
In this case, there are 13 and 10 bulges with $n_{\rm sph}<2$ that lie above and below, respectively, the correlation  
traced by $n_{\rm sph}>3$ bulges.}, 
and that the amplitude of their offset is the same ($\lesssim 1.5\rm~dex$).
That is, bulges with $n_{\rm sph}<2$ do not appear to be offset from the correlation traced by bulges with $n_{\rm sph}>2$.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\columnwidth]{images/inset_psb.pdf}
\caption{Vertical offset from the $M_{\rm BH} - L_{\rm sph}$ correlation defined by spheroids with $n_{\rm sph} \geq 2$, 
plotted against $n_{\rm sph}$. 
The vertical dashed line corresponds to $n_{\rm sph} = 2$.
The horizontal solid line is equivalent to a zero vertical offset.
Among bulges with $n_{\rm sph}<2$, 12 have positive vertical offset and 11 have negative vertical offset.
Hence, bulges with $n_{\rm sph}<2$ are not randomly offset to lower black hole masses 
from the correlation traced by bulges with $n_{\rm sph} \geq 2$.}
\label{fig:pseudob}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\columnwidth]{images/mbh_vs_mass_sph.pdf}
\caption{Black hole mass plotted against spheroid stellar mass. 
Symbols are coded according to the galaxy morphological type.
The red dashed line indicates the BCES bisector linear regression for the bulges of early-type galaxies (E+S0), 
with the red shaded area denoting its $1\sigma$ uncertainty. 
The bulges of early-type galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$,
a near-linear relation consistent with a dry-merging formation scenario.
The blue solid line shows the BCES bisector linear regression for the bulges of late-type (Sp) galaxies, 
with the blue shaded area denoting its $1\sigma$ uncertainty. 
The bulges of late-type galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently (``quadratically'' or ``cubically'') than the host bulge. }
\label{fig:mbhmasssph}
\end{center}
\end{figure}

@@

\section{Conclusions}
\label{sec:concl}
%ngc0524 outlier, n1332 non outlier
to be re-written

\begin{itemize}
\item The bulges of early-type (E + S0) galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$, 
consistent with a dry-merging formation scenario,
and define a tight \emph{red sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$.
On the other hand, the bulges of late-type (Sp) galaxies define a much steeper \emph{blue sequence}, 
with $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently than the host bulge. 
\item S\'ersic galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{x \pm x}$, a less steep sequence than previously reported; 
\item bulges with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
are not offset to lower $M_{\rm BH}$ from the correlation defined by bulges with $n_{\rm sph}>2$; 
\item $L_{\rm sph}$ and $L_{\rm gal}$ correlate equally well with $M_{\rm BH}$, in terms of intrinsic scatter, only for early-type galaxies; 
once reasonable numbers of spiral galaxies are included, the correlation with $L_{\rm sph}$ is better than that with $L_{\rm gal}$.
\end{itemize}

>>>>

\begin{abstract}
{\bf limit of 250 words, currently there are 229 words, there's room for 21 words. 
Things not mentioned: 1) the galaxy components (more appropriate for data paper?????), 
2) $3.6\rm~\mu m$ \emph{Spitzer} satellite imagery, minimally affected by dust and star formation 
3) updated linear regressions (symmetrical, to be compared with theoretical expectations, and non-symmetrical, to be used to predict black hole masses) 
for S\'ersic/core-S\'ersic galaxies and for different morphological types (E/S0, Sp)}\\
We present correlations between black hole mass, $M_{\rm BH}$, 
and both the host galaxy's and spheroid's infrared luminosity, $L_{\rm gal}$ and $L_{\rm sph}$, 
and the spheroid's stellar mass, $M_{\rm *,sph}$,
built upon a high-quality data set.
Our sample, the largest ever used, counts 66 local galaxies with a dynamical measurement of $M_{\rm BH}$.
%Luminosities come from high signal-to-noise $3.6\rm~\mu m$ \emph{Spitzer} satellite imagery, 
%which is minimally affected by dust and star formation. 
%These luminosities were derived from our state-of-the-art multicomponent galaxy decompositions, 
%that take into account bulges, disks, bars, spiral arms, rings, haloes, extended or unresolved nuclear sources and partially depleted cores, 
%and that -- for the first time -- were checked to be consistent with the galaxy kinematics. 
Spheroid luminosities come from high signal-to-noise $3.6\rm~\mu m$ \emph{Spitzer} satellite imagery, 
and were derived from our state-of-the-art {\bf ``many-component''} galaxy decompositions 
that, for the first time, were checked to be consistent with the galaxy kinematics. 
Previous studies have used galaxy samples that were overwhelmingly dominated by high-mass, early-type objects.
Instead, our sample includes 17 spiral galaxies, half of which have $M_{\rm BH} < 10^7\rm~M_\odot$, 
and allows us to better investigate the poorly studied low-mass end of the $M_{\rm BH} - M_{\rm *,sph}$ correlation.
%We provide updated linear regressions 
%(symmetrical, to be compared with theoretical expectations, and non-symmetrical, to be used to predict black hole masses) 
%for S\'ersic/core-S\'ersic galaxies and for different morphological types (E/S0, Sp). 
The bulges of early-type (E + S0) galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$, 
consistent with a dry-merging formation scenario,
and define a tight \emph{red sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$.
On the other hand, the bulges of late-type (Sp) galaxies define a much steeper \emph{blue sequence}, 
with $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently than the host bulge. 
We additionally report that: i) S\'ersic galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.48 \pm 0.20}$, a less steep sequence than previously reported; 
ii) bulges with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
are not offset to lower $M_{\rm BH}$ from the correlation defined by bulges with $n_{\rm sph}>2$; 
iii) $L_{\rm sph}$ and $L_{\rm gal}$ correlate equally well with $M_{\rm BH}$, in terms of intrinsic scatter, only for early-type galaxies; 
once reasonable numbers of spiral galaxies are included, the correlation with $L_{\rm sph}$ is better than that with $L_{\rm gal}$.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\columnwidth]{images/mbh_vs_mag_tot.pdf}
\caption{Black hole mass plotted against $3.6\rm~\mu m$ galaxy absolute magnitude. 
Symbols are coded according to the galaxy morphological type: red circle = E, red star = E/S0, 
red upward triangle = S0, blue downward triangle = S0/Sp, blue square = Sp, black diamond = merger. 
Empty symbols represent core-S\'ersic spheroids, whereas filled symbols are used for S\'ersic spheroids. 
Four spiral galaxies had their magnitudes overestimated and are shown as upper limits. 
The red dashed line indicates the BCES bisector linear regression for 45 early-type galaxies (E+S0), 
with the red shaded area denoting its $1\sigma$ uncertainty. 
We were not able to obtain any meaningful linear regression for the late-type (Sp) galaxies. 
$M_{\rm BH}$ correlates equally well with $L_{\rm sph}$ and $L_{\rm gal}$ only for early-type galaxies, 
but not for all (early+late) galaxies. }
\label{fig:mbhmaggal}
\end{center}
\end{figure}

@@

\subsection{Pseudo- versus classical bulges}
Current views distinguish between classical bulges, which are considered to be spheroidal, pressure-supported systems, 
formed through violent processes, such as hierarchical clustering via minor mergers, 
and pseudo-bulges, thought to be disk-like, rotation-supported systems, 
built from secular evolution processes, such as instabilities of their surrounding disk or bar. 
Pseudo-bulges are notoriously hard to identify \citep{graham2013review,graham2014review,graham2015pseudo,graham2015review}.
For example, mergers can create bulges that rotate (e.g.~\citealt{bekki2010,keselmannusser2012}), 
and bars can spin-up classical bulges (e.g.~\citealt{saha2012}), 
thus rotation is not a definitive signature of a pseudobulge. 
Furthermore, many galaxies host both a pseudo- and a classical bulge (e.g.~\citealt{erwin2003,erwin2015,athanassoula2005,Gadotti2009,
macarthur2009,dosanjosdasilva2013,seidel2015}). 
In the recent literature, pseudo- and classical bulges have been frequently told apart by means of a divide at a 
S\'ersic index of $n_{\rm sph}=2$ (e.g.~\citealt{sani2011,beifiori2012}), 
although, from a selection of hundreds of disc galaxies imaged in the $K$-band, 
\cite{grahamworley2008} observed no such bimodality in the bulge S\'ersic indices. 
While pseudo-bulges are expected to have exponential-like surface brightness profiles ($n_{\rm sph} \simeq 1$), 
being disky components that formed from their surrounding exponential disks 
(e.g.~\citealt{bardeen1975,hohl1975,combessanders1981,combes1990,pfennigerfriedli1991}), 
it has been shown that mergers can create bulges with $n_{\rm sph}<2$
(e.g.~\citealt{elichemoral2011,scannapieco2011,querejeta2015}), 
just as low-luminosity elliptical galaxies (not built from the secular evolution of a disk)
are well known to have $n_{\rm sph}<2$ and even $n_{\rm sph}<1$ (e.g.~\citealt{davies1988,youngcurrie1994}). \\
\cite{sani2011} reported that pseudo-bulges -- which they labelled as such according to the $n_{\rm sph}<2$ criterion -- 
with low black hole masses ($M_{\rm BH} < 10^7\rm~M_\odot$) are significantly displaced from the correlation 
traced by classical bulges. 
In Figure \ref{fig:pseudob}, we show the distribution of spheroid S\'ersic indices\footnote{The spheroid S\'ersic index 
is taken from our one-dimensional fit of the galaxy's major-axis surface brightness profile (\emph{Paper I}).} 
in the $M_{\rm BH} - L_{\rm sph}$ diagram.
Our aim is to check whether $n_{\rm sph}<2$ bulges 
are offset to lower black hole masses from the correlation defined by $n_{\rm sph}>2$ bulges. 
To do this, we fit a symmetrical linear regression to the bulges that have $n_{\rm sph}>2$ 
and we compute the vertical offset of all bulges from the regression. 
In the inset of Figure \ref{fig:pseudob}, we plot the vertical offset against $n_{\rm sph}$. 
Among the 23 bulges with $n_{\rm sph}<2$, 12 have positive vertical offset and 11 have negative vertical offset. 
\cite{kormendy2015review} provides a list of many pseudo-bulges classification criteria, including the divide at $n_{\rm sph}=2$, 
and cautions that each individual criterion has a failure rate of 0-25\%. 
If this is true, we should have that no less than 75\% of $n_{\rm sph}<2$ bulges display negative vertical offset. 
What we observe, instead, is that there are the same number of $n_{\rm sph}<2$ bulges lying above and below 
the correlation defined by $n_{\rm sph}>2$ bulges\footnote{One 
reaches the same conclusion when using the vertical offset from the correlation defined by bulges with $n_{\rm sph}>3$. 
In this case, there are 13 and 10 bulges with $n_{\rm sph}<2$ that lie above and below, respectively, the correlation  
traced by $n_{\rm sph}>3$ bulges.}, 
and that the amplitude of their offset is the same ($\lesssim 1.5\rm~dex$).
That is, bulges with $n_{\rm sph}<2$ do not appear to be offset from the correlation traced by bulges with $n_{\rm sph}>2$.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\columnwidth]{images/inset_psb.pdf}
\caption{Vertical offset from the $M_{\rm BH} - L_{\rm sph}$ correlation defined by spheroids with $n_{\rm sph} \geq 2$, 
plotted against $n_{\rm sph}$. 
The vertical dashed line corresponds to $n_{\rm sph} = 2$.
The horizontal solid line is equivalent to a zero vertical offset.
Among the bulges with $n_{\rm sph}<2$, 12 have positive vertical offset and 11 have negative vertical offset.
Hence, bulges with $n_{\rm sph}<2$ are not randomly offset to lower black hole masses 
from the correlation traced by bulges with $n_{\rm sph} \geq 2$.}
\label{fig:pseudob}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\columnwidth]{images/mbh_vs_mass_sph.pdf}
\caption{Black hole mass plotted against spheroid stellar mass. 
Symbols are coded according to the galaxy morphological type.
The red dashed line indicates the BCES bisector linear regression for the bulges of early-type galaxies (E+S0), 
with the red shaded area denoting its $1\sigma$ uncertainty. 
The bulges of early-type galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$,
a near-linear relation consistent with a dry-merging formation scenario.
The blue solid line shows the BCES bisector linear regression for the bulges of late-type (Sp) galaxies, 
with the blue shaded area denoting its $1\sigma$ uncertainty. 
The bulges of late-type galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
indicating that gas-rich processes feed the black hole more efficiently (``quadratically'' or ``cubically'') than the host bulge. 
We note that AGNs with $10^5 \lesssim M_{\rm BH}/{\rm M_\odot} \lesssim 2 \times 10^6$ appear to roughly follow the blue line 
(see Graham et al. 2015, \emph{in preparation}).}
\label{fig:mbhmasssph}
\end{center}
\end{figure}

@@

\section{Conclusions}
\label{sec:concl}
% %ngc0524 outlier, n1332 non outlier
% to be re-written
%
% \begin{itemize}
% \item The bulges of early-type (E + S0) galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{1.04 \pm 0.10}$, 
% consistent with a dry-merging formation scenario,
% and define a tight \emph{red sequence} with intrinsic scatter $\epsilon_{(Y|X)} = 0.43 \pm 0.06\rm~dex$.
% On the other hand, the bulges of late-type (Sp) galaxies define a much steeper \emph{blue sequence}, 
% with $M_{\rm BH} \propto M_{\rm *,sph}^{2-3}$, 
% indicating that gas-rich processes feed the black hole more efficiently than the host bulge. 
% \item S\'ersic galaxies follow $M_{\rm BH} \propto M_{\rm *,sph}^{x \pm x}$, a less steep sequence than previously reported; 
% \item bulges with S\'ersic index $n_{\rm sph}<2$, argued by some to be pseudo-bulges, 
% are not offset to lower $M_{\rm BH}$ from the correlation defined by bulges with $n_{\rm sph}>2$; 
% \item $L_{\rm sph}$ and $L_{\rm gal}$ correlate equally well with $M_{\rm BH}$, in terms of intrinsic scatter, only for early-type galaxies; 
% once reasonable numbers of spiral galaxies are included, the correlation with $L_{\rm sph}$ is better than that with $L_{\rm gal}$.
% \end{itemize}
